---
output:
  BiocStyle::html_document
---

<!---
Because we split the analysis pipeline in different independent files,
to speed up processing it, here in the setup block we load libraries and
objects that were loaded or produced in the previously processed file,
and which are necessary in this file.
--->

```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)
#setwd("/home/guille/Guille/MBHS/3T/IEO/project")

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

se <- readRDS("seKIRC.rds")
#se <- readRDS(file.path("rawCounts", "seKICH.rds"))
#dge <- readRDS(file.path("results", "dge.rds"))
se.filt.unnorm <- readRDS(file.path("results", "se.filt.unnorm.rds"))
dge.filt.unnorm <- readRDS(file.path("results", "dge.filt.unnorm.rds"))
se.filt <- readRDS(file.path("results", "se.filt.rds"))
dge.filt <- readRDS(file.path("results", "dge.filt.rds"))

```

## Differential expression

# fold-change plot

```{r, message=FALSE}

dim(se.filt)

logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)
normalExp <- rowMeans(logCPM[, se.filt$type == "normal"])
tumorExp <- rowMeans(logCPM[, se.filt$type == "tumor"])

par(mfrow = c(1, 2))
plot(normalExp, tumorExp, xlab = "Normal", ylab = "Tumor", pch = ".", cex = 4, las = 1)
plot((tumorExp + normalExp)/2, tumorExp - normalExp, pch = ".", cex = 4, las = 1)
```

# Fold-change significant genes

```{r, message=FALSE}
log2fc <- tumorExp - normalExp
ranking <- order(abs(log2fc), decreasing = TRUE)
fc_genes <- data.frame(Log2FC = round(log2fc[ranking], digits = 3), FC = round(2^log2fc[ranking],
digits = 3), `1/FC` = round(2^(-log2fc[ranking]), digits = 3), row.names = rowData(se)$symbol[ranking], check.names = FALSE)
head(fc_genes)
```

# SVA adjustment

We perform a simple examination of expression changes and their associated p-values
using the R/Bioconductor package [sva](http://bioconductor.org/packages/sva).

```{r, message=FALSE}
library(sva)
mod <- model.matrix(~ se.filt$type, colData(se.filt))
mod0 <- model.matrix(~ 1, colData(se.filt))
pv <- f.pvalue(assays(se.filt)$logCPM, mod, mod0)
sum(p.adjust(pv, method="fdr") < 0.01)
```

There are `r sum(p.adjust(pv, method="fdr") < 0.01)` genes changing significantly
their expression at FDR < 1%. In Figure \@ref(fig:pdist) below we show the distribution of the
resulting p-values.

```{r pdist, echo=FALSE, out.width="400px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples."}
hist(pv, main="", las=1)
```

Now, let's estimate surrogate variables using the `sva()` function.

```{r}
sv <- sva(assays(se.filt)$logCPM, mod, mod0)
sv$n
```

The SVA algorithm has found `r sv$n` surrogate variables. Let's use them to
assess againt the extent of differential expression this time adjusting for these
surrogate variables.

```{r}
modsv <- cbind(mod, sv$sv)
mod0sv <- cbind(mod0, sv$sv)
pvsv <- f.pvalue(assays(se.filt)$logCPM, modsv, mod0sv)
sum(p.adjust(pvsv, method="fdr") < 0.01)
```

We have increased the number of changing genes to `r sum(p.adjust(pvsv, method="fdr") < 0.01)`.
Figure \@ref(fig:psvdist) shows the resulting distribution of p-values.

```{r psvdist, echo=FALSE, out.width="400px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples, adjusting for surrogate variables estimated with SVA."}
hist(pvsv, main="", las=1)
```

## Session information

```{r, message=FALSE}
sessionInfo()
```
